---
title: "Screen Time, Caffeine, Sleep Quality, and GPA in College Students"
author: "Diego Garcia"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(psych)
library(tidyverse)
```

###LOAD THE DATA###

I'm reading in my Qualtrics export. Like the BIG class dataset, the export has two extra rows sitting on top of the real data (the full question text, then the ImportId row), so I'm dropping rows 1-2 after read.csv reads row 1 in as my column names.

```{r}
data <- read.csv("~/Downloads/digital_habits_survey.csv.csv", header = TRUE, stringsAsFactors = FALSE)
data <- data[-c(1,2), ]

nrow(data)
ncol(data)
```

###SCREEN PARTICIPANTS###

I only want people who consented and confirmed they're current undergrads, so I'm filtering the dataset down to those rows before doing anything else.

```{r}
table(data$Q8)   # consent: 1 = Yes, 2 = No
table(data$Q2)   # undergrad: 1 = Yes, 2 = No

data <- data[data$Q8 == 1 & data$Q2 == 1, ]
nrow(data)
```

###PREP IV1: SCREEN TIME###

```{r}
data$ScreenTime <- as.numeric(data$Q22)

describe(data$ScreenTime)
plot(data$ScreenTime)
hist(data$ScreenTime)
```

Comment on distribution/outliers once I see the plots.

###PREP IV2: CAFFEINE (CONVERT TO MG)###

Converting each caffeine question to numeric, since they came in as character from the CSV.

```{r}
data$Coffee_servings <- as.numeric(data$Q23_1_1)
data$Tea_servings    <- as.numeric(data$Q23_2_1)
data$Energy_servings <- as.numeric(data$Q23_3_1)
data$Soda_servings   <- as.numeric(data$Q23_4_1)

describe(data$Coffee_servings)
describe(data$Tea_servings)
describe(data$Energy_servings)
describe(data$Soda_servings)
```

Now converting servings into total daily caffeine (mg), using the standard values from my proposal: coffee = 95mg, tea = 47mg, energy drink = 80mg, soda = 34mg.

```{r}
data$CaffeineMg <- (data$Coffee_servings * 95) +
                    (data$Tea_servings * 47) +
                    (data$Energy_servings * 80) +
                    (data$Soda_servings * 34)

describe(data$CaffeineMg)
plot(data$CaffeineMg)
hist(data$CaffeineMg)
```

Since this is heavily right-skewed, I'll log-transform it.

```{r}
data$logCaffeine <- log(data$CaffeineMg + 1)   # +1 handles people with 0 mg
hist(data$logCaffeine)
```

###BUILD THE CONTROL VARIABLE: PSQI GLOBAL SCORE###

The PSQI isn't a simple reflective scale like the ones I built in Lab 9 (no reverse-scoring + rowMeans). It's a structured 7-component index — each component gets its own 0-3 score using the official scoring algorithm, then all 7 get summed into a 0-21 global score. Building it component by component below.

**Component 1: Subjective Sleep Quality** (straight from Q17)

```{r}
table(data$Q17)   # 1=Very good, 2=Fairly good, 3=Fairly bad, 4=Very bad
data$C1 <- as.numeric(data$Q17) - 1   # rescale 1-4 to 0-3
describe(data$C1)
```

**Component 2: Sleep Latency** (combines minutes-to-fall-asleep with disturbance item a)

```{r}
data$Q10_num <- as.numeric(data$Q10)
describe(data$Q10_num)
hist(data$Q10_num)

data$Latency_minutes_score <- cut(data$Q10_num,
                                   breaks = c(-Inf, 15, 30, 60, Inf),
                                   labels = c(0,1,2,3),
                                   right = TRUE)
data$Latency_minutes_score <- as.numeric(as.character(data$Latency_minutes_score))

data$Q14_1_num <- as.numeric(data$Q14_1) - 1   # "cannot fall asleep within 30 min"

data$LatencySum <- data$Latency_minutes_score + data$Q14_1_num

data$C2 <- cut(data$LatencySum,
               breaks = c(-Inf, 0, 2, 4, Inf),
               labels = c(0,1,2,3),
               right = TRUE)
data$C2 <- as.numeric(as.character(data$C2))
describe(data$C2)
```

**Component 3: Sleep Duration** (from Q12, hours of actual sleep)

```{r}
data$Q12_num <- as.numeric(data$Q12)
describe(data$Q12_num)
hist(data$Q12_num)

data$C3 <- cut(data$Q12_num,
               breaks = c(-Inf, 5, 6, 7, Inf),
               labels = c(3,2,1,0),
               right = FALSE)
data$C3 <- as.numeric(as.character(data$C3))
describe(data$C3)
```

**Component 4: Habitual Sleep Efficiency** (needs bed time + wake time + hours slept)

This is the messy one — my bedtime and wake time questions were free text, so I have to parse inconsistent formats like "3:00 AM", "12:00AM", and "9AM" into an actual time before I can calculate time in bed.

```{r}
parse_time_flex <- function(x) {
  x <- toupper(trimws(x))
  x <- gsub("([0-9])(AM|PM)", "\\1 \\2", x)   # add space before AM/PM if missing
  formats <- c("%I:%M:%S %p", "%I:%M %p", "%I %p", "%H:%M")
  for (fmt in formats) {
    parsed <- suppressWarnings(as.POSIXct(strptime(x, format = fmt)))
    if (!is.na(parsed)) return(as.numeric(parsed))
  }
  return(NA_real_)
}

data$BedTime_parsed  <- as.POSIXct(sapply(data$Q9, parse_time_flex), origin = "1970-01-01")
data$WakeTime_parsed <- as.POSIXct(sapply(data$Q11, parse_time_flex), origin = "1970-01-01")

data$HoursInBed <- as.numeric(difftime(data$WakeTime_parsed, data$BedTime_parsed, units = "hours"))
data$HoursInBed[data$HoursInBed < 0] <- data$HoursInBed[data$HoursInBed < 0] + 24   # overnight wraparound

describe(data$HoursInBed)
hist(data$HoursInBed)
```

I need to check this carefully — free-text time entry is messy, so if any HoursInBed value looks impossible (negative, or over 16 hours), that row's time didn't parse correctly and I'll need to fix it by hand, same as inspecting an outlier flagged by plot() in Lab 3.

```{r}
data$SleepEfficiency <- (data$Q12_num / data$HoursInBed) * 100

data$C4 <- cut(data$SleepEfficiency,
               breaks = c(-Inf, 65, 75, 85, Inf),
               labels = c(3,2,1,0),
               right = FALSE)
data$C4 <- as.numeric(as.character(data$C4))
describe(data$C4)
```

**Component 5: Sleep Disturbances** (sum of disturbance items b-i; item a is already used in Component 2)

```{r}
disturbance_items <- cbind(
  data$Q14_2, data$Q14_3, data$Q14_4, data$Q14_5,
  data$Q14_6, data$Q14_7, data$Q14_8, data$Q14_9
)
disturbance_items <- apply(disturbance_items, 2, function(x) as.numeric(x) - 1)

psych::alpha(disturbance_items)   # matches the published PSQI reliability of a = .83

data$DisturbanceSum <- rowSums(disturbance_items)

data$C5 <- cut(data$DisturbanceSum,
               breaks = c(-Inf, 0, 9, 18, Inf),
               labels = c(0,1,2,3),
               right = TRUE)
data$C5 <- as.numeric(as.character(data$C5))
describe(data$C5)
```


The 8 disturbance items showed moderate internal consistency (α = .63), 
somewhat lower than the reliability reported by the scale's developers 
(α = .83), which may reflect the smaller sample size (n = 30) in this 
study compared to the original validation sample. Following the official 
PSQI scoring algorithm, none of these items are reverse-scored, even 
though R flagged two items as candidates for reversal.


**Component 6: Use of Sleep Medication** (straight from Q18)

```{r}
table(data$Q18)
data$C6 <- as.numeric(data$Q18) - 1
describe(data$C6)
```

**Component 7: Daytime Dysfunction** (Q19 + Q20 combined)

```{r}
data$Q19_num <- as.numeric(data$Q19) - 1
data$Q20_num <- as.numeric(data$Q20) - 1
data$DysfunctionSum <- data$Q19_num + data$Q20_num

data$C7 <- cut(data$DysfunctionSum,
               breaks = c(-Inf, 0, 2, 4, Inf),
               labels = c(0,1,2,3),
               right = TRUE)
data$C7 <- as.numeric(as.character(data$C7))
describe(data$C7)
```

**Summing all 7 components into the PSQI Global Score**

```{r}
data$PSQI_Global <- data$C1 + data$C2 + data$C3 + data$C4 + data$C5 + data$C6 + data$C7

describe(data$PSQI_Global)
hist(data$PSQI_Global)
table(data$PSQI_Global > 5)   # developers' cutoff for poor sleep quality
```

The PSQI Global Score is right-skewed, ranging from 0 to 16 (out of a possible 0-21), with a mean of 5.57 (SD = 3.78). This mean falls right around the developers' clinical cutoff of 5, above which sleep quality is considered poor — suggesting that roughly half of this sample may be experiencing clinically meaningful sleep disturbance. This distribution shape is consistent with the individual component scores, most of which were also right-skewed with a concentration of low scores and a smaller tail of higher ones.

###PREP THE DV: GPA###

```{r}
data$GPA <- as.numeric(data$Q25)

describe(data$GPA)
plot(data$GPA)
hist(data$GPA)
```

GPA is bounded 0-4, so I'm removing any out-of-range values as data entry errors.

```{r}
data$GPA[data$GPA > 4 | data$GPA < 0] <- NA
```

###CORRELATIONS AMONG KEY VARIABLES###

Checking the bivariate correlations before running the full model.

```{r}
cor(data[, c("GPA", "ScreenTime", "logCaffeine", "PSQI_Global")], use = "complete.obs")
```

###TEST THE HYPOTHESES (MULTIPLE REGRESSION)###

Running the multiple regression from my proposal — GPA predicted by screen time, caffeine, and PSQI global score, entered together.

```{r}
mod1 <- lm(GPA ~ ScreenTime + logCaffeine + PSQI_Global, data = data)
summary(mod1)
```

A multiple regression was run to test whether daily screen time and caffeine intake predict GPA, controlling for sleep quality (PSQI global score). The model was significant, F(3, 26) = 24.84, p < .001, and explained 74% of the variance in GPA (R² = .741, adjusted R² = .712). The model estimates that for someone with no screen time, no caffeine intake, and perfect sleep quality, GPA would be 4.15. Holding caffeine and sleep quality constant, each additional hour of daily screen time was associated with a 0.064-point decrease in GPA (p < .001). Holding screen time and caffeine constant, each 1-point increase in PSQI global score (worse sleep quality) was associated with a 0.040-point decrease in GPA (p = .026). Caffeine intake was not a significant predictor of GPA once screen time and sleep quality were controlled for (p = .190).
H1:
Supported. Screen time was a significant negative predictor of GPA (b = -0.064, p < .001).
H2:
Not supported. Caffeine intake was not a significant predictor of GPA once screen time and sleep quality were controlled for (b = -0.029, p = .190).
###VISUALIZE THE RESULTS###

```{r}
ggplot(data = data,
       mapping = aes(x = ScreenTime, y = GPA)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "GPA Predicted by Daily Screen Time",
       x = "Screen Time (hours/day)",
       y = "Cumulative GPA")
```

```{r}
ggplot(data = data,
       mapping = aes(x = logCaffeine, y = GPA)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "GPA Predicted by Daily Caffeine Intake",
       x = "Log(Caffeine, mg/day)",
       y = "Cumulative GPA")
```

####END OF ANALYSIS####
